RationalGRL: A Framework for Argumentation and Goal Modeling

### 2017-11-07 formalframework.tex
small fixes

The definition above is sufficient to capture all intentional elements (IEs) used in GRL. %In the next definition, we present some simplifying notational conventions.
%\begin{definition}[Notation]
%\label{def:notation}
%We adopt three conventions simplifying our notation:
%\begin{itemize}
%\item
%We refer to the element of a tuple using the dot (".") notation. That is, we may for instance refer to the id, type, name and decomposition type of an $IE$ with respectively $ie.id$, $ie.type$, $ie.name$, and $ie.decomptype$.
%\item 
%We refer to a set of elements with the same id $i$  using the $i$ subscript on the set. For instance a set of IEs with id $i$ is denoted by $IE_i$, and if this is a single element, we denote it by $ie_i$\footnote{In GRL, there always exists at most one element for every id (see Def.~\ref{def:grl-model}), in RationalGRL, however, this condition does not hold (see Def.~\ref{def:rationalgrl-model}).} For instance, we may refer to the intentional element $ie = (0, Goal, \text{Make profit}, AND)$ with $ie_0$ and write $ie_0.type = Goal$, $ie_0.name = \text{Make profit}$, and $ie_0.decompositiontype = AND$. 
%\item
%We can also refer to intentional elements of a specific type simply by $type_{id}$. For instance, we can abbreviate the element in the previous item with $goal_0$ and write $goal_0.name = \text{Make profit}$, and $goal_0.decomptype = AND$.
%\end{itemize}
%\end{definition}
Throughout this section we use the example in Figure~\ref{fig:example-small}. This example contains various IEs which can be formalized using Definition~\ref{def:ie}, for example, 
$(1, Resource, \text{Car objects}, AND)$, $(2, Softgoal,$ Realistic simulation$, AND)$ and $(5, Goal,$ Generate traffic$, XOR)$.

%Using the short-hand notation defined in Def.~\ref{def:notation}, we can make the following statements:
%\begin{itemize}
%\item $goal_4.name = \text{Generate traffic}$
%\item $goal_4.decomptype = XOR$
%\item $resource_1.name = \text{Car objects}$
%\end{itemize}
%\begin{figure}[ht]
%\centering
%\includegraphics[width=\columnwidth]{img/Example1-new.pdf}
%\caption{Example GRL model (reprint of Fig.~\ref{fig:example-small})}
%\label{fig:example-small2}
%\end{figure} 
We now define actors.

@@

\begin{flalign*}
&IE=&\{&(1, Resource, \text{Car objects}, AND),&\\
&   &  &(2, Softgoal, \text{Realistic simulation}, AND),&\\
&   &  &(3, Softgoal, \text{Simple design}, AND),&\\
&   &  &(4, Softgoal, \text{Easy to use}, AND),&\\
&   &  &(5, Goal, \text{Generate traffic}, XOR),&\\
&   &  &(6, Task, \text{Create new cars}, AND),&\\
&   &  &(7, Task, \text{Keep same cars}, AND)\}&\\
&Act=& &\{(0, Actor, \text{Traffic Simulator})\}&\\
&R_{ActIE}=& &\{(ActIE, 0, 1), \ldots, (ActIE,0,7)\}&\\
&Link=&\{&(8, Dep, 4, 1)&\\
&     & &(9, PosContr, 6, 2)&\\
&     & &(10, NegContr, 6, 3)&\\
&     & &(11, Decomp, 6, 5)&\\
&     & &(12, Decomp, 7, 5)\}&\\
\end{flalign*}

@@

Examples of arguments in Figure~\ref{fig:example-small3} are $A_5 = (5,$ $Goal,$ Generate traffic$, AND)$, $A'_5 = (5, Goal,$ Generate traffic$, XOR)$, $A_4 = (4, Softgoal,$ Easy to use$, AND)$ and $A_{13}=(13,$ $ArgElem,$ Redundant$)$. Note that the two arguments $A_5$ and $A'_5$ show an important difference between RationalGRL models and valid GRL models: While a valid GRL model disallows multiple elements with the same identifier (Definition~\ref{def:valid-grl-model}, condition 1), RationalGRL models do not enforce this restriction. This is because it is possible to create multiple arguments for the same element, where each argument contains different content for the same element. However, the set of \emph{accepted} elements in a RationalGRL should all have unique identifier (see Definition~\ref{def:valid-rationalgrl-model}).

@@

As an example, take the arguments $A_5$ for the goal `Generate traffic' ($AND$ decomposition), $A'_5$ for the goal `Generate traffic' ($XOR$ decomposition), $A_4$ for the softgoal `Easy to use' and the generic argument $A_{13}$ (`Redundant') against the softgoal (`Easy to use'). Given these arguments there are two attack links (Figure~\ref{fig:example-small3}), namely $(A'_{5},A_{5})$ and $(A_{13}, A_{4})$.

@@

\begin{flalign*}
A_0 = &(0, Actor, \text{Traffic simulator})&\\
A_1 = &(1, Task, \text{Car objects}, AND),&\\
A_2 = &(2, Softgoal, \text{Realistic simulation}, AND),&\\
A_3 = &(3, Softgoal, \text{Simple design}, AND),&\\
A_4 = &(4, Softgoal, \text{Easy to use}, AND),&\\
A_5 = &(5, Goal, \text{Generate traffic}, AND),&\\
A'_5 = &(5, Goal, \text{Generate traffic}, XOR),&\\
A_6 = &(6, Task, \text{Create new cars}, AND),&\\
A_7 = &(7, Task, \text{Wrap cars}, AND),&\\
A'_7 = &(7, Task, \text{Keep same cars}, AND),&\\
A_8 = &(8, Dep, 4, 1),&\\
A_9 = &(9, PosContr, 6, 2),&\\
A_{10} = &(10, NegContr, 6, 3),&\\
A_{11} = &(11, Decomp, 6, 5),&\\
A_{12} = &(12, Decomp, 7, 5),&\\
A_{13} = &(13, GenArg, \text{Redundant}),&\\
A_{14} = &(14, GenArg, \text{Necessary}),&\\
A_{15} = &(15, GenArg, \text{No effect}),&\\
A_{16} = &(16, GenArg, \text{Changes traffic flow}),&\\
A_{17} = &(ActIE,0,1),\ldots,A_{23}=(ActIE,0,7)&\\
\end{flalign*}
This model is then formalized as $RatGRL=(Arg, Att)$ where:
\begin{flalign*}
Arg = &\{A_0,\ldots,A_5, A'_5,A_6,A_7,A'_7\ldots,A_{23}\}\\
Att= &\{(A'_{5},A_5), (A'_{7},A_{7}), (A_{13},A_{4}), (A_{14},A_{13}),\\
     & (A_{15},A_{9}),(A_{16},A_{15})\}\\
\end{flalign*}

@@

As an example, suppose we are adding the decomposition of goal ``Generate traffic'', expressed as argument $A_5 = (5,$ $Goal,$ Generate traffic$, AND)$, into task ``Keep same cars'', and suppose we have the following argument for the goal: . Adding the decomposition can be formalized as executing $AS_5(5, \text{Keep same cars})$, and results in two new arguments: $A_7 = (7, Task, \text{Keep same cars}, AND)$ and $A_{12} = (12, Decomp, 7, 5)$, which are both added to the set of arguments $Arg$.

>>>>

The definition above is sufficient to capture all intentional elements (IEs) used in GRL. %In the next definition, we present some simplifying notational conventions.
%\begin{definition}[Notation]
%\label{def:notation}
%We adopt three conventions simplifying our notation:
%\begin{itemize}
%\item
%We refer to the element of a tuple using the dot (".") notation. That is, we may for instance refer to the id, type, name and decomposition type of an $IE$ with respectively $ie.id$, $ie.type$, $ie.name$, and $ie.decomptype$.
%\item 
%We refer to a set of elements with the same id $i$  using the $i$ subscript on the set. For instance a set of IEs with id $i$ is denoted by $IE_i$, and if this is a single element, we denote it by $ie_i$\footnote{In GRL, there always exists at most one element for every id (see Def.~\ref{def:grl-model}), in RationalGRL, however, this condition does not hold (see Def.~\ref{def:rationalgrl-model}).} For instance, we may refer to the intentional element $ie = (0, Goal, \text{Make profit}, AND)$ with $ie_0$ and write $ie_0.type = Goal$, $ie_0.name = \text{Make profit}$, and $ie_0.decompositiontype = AND$. 
%\item
%We can also refer to intentional elements of a specific type simply by $type_{id}$. For instance, we can abbreviate the element in the previous item with $goal_0$ and write $goal_0.name = \text{Make profit}$, and $goal_0.decomptype = AND$.
%\end{itemize}
%\end{definition}
Throughout this section we use the example in Figure~\ref{fig:example-small}. This example contains various IEs which can be formalized using Definition~\ref{def:ie}, for example, 
$(1, Resource, \text{Car objects}, AND)$, $(2, Softgoal,$ Realistic simulation$, AND)$ and $(5, Goal,$ Generate cars$, XOR)$.

%Using the short-hand notation defined in Def.~\ref{def:notation}, we can make the following statements:
%\begin{itemize}
%\item $goal_4.name = \text{Generate cars}$
%\item $goal_4.decomptype = XOR$
%\item $resource_1.name = \text{Car objects}$
%\end{itemize}
%\begin{figure}[ht]
%\centering
%\includegraphics[width=\columnwidth]{img/Example1-new.pdf}
%\caption{Example GRL model (reprint of Fig.~\ref{fig:example-small})}
%\label{fig:example-small2}
%\end{figure} 
We now define actors.

@@

\begin{flalign*}
&IE=&\{&(1, Resource, \text{Car objects}, AND),&\\
&   &  &(2, Softgoal, \text{Realistic simulation}, AND),&\\
&   &  &(3, Softgoal, \text{Simple design}, AND),&\\
&   &  &(4, Softgoal, \text{Easy to use}, AND),&\\
&   &  &(5, Goal, \text{Generate cars}, XOR),&\\
&   &  &(6, Task, \text{Create new cars}, AND),&\\
&   &  &(7, Task, \text{Keep same cars}, AND)\}&\\
&Act=& &\{(0, Actor, \text{Traffic Simulator})\}&\\
&R_{ActIE}=& &\{(ActIE, 0, 1), \ldots, (ActIE,0,7)\}&\\
&Link=&\{&(8, Dep, 4, 1)&\\
&     & &(9, PosContr, 6, 2)&\\
&     & &(10, NegContr, 6, 3)&\\
&     & &(11, Decomp, 6, 5)&\\
&     & &(12, Decomp, 7, 5)\}&\\
\end{flalign*}

@@

Examples of arguments in Figure~\ref{fig:example-small3} are $A_5 = (5,$ $Goal,$ Generate cars$, AND)$, $A'_5 = (5, Goal,$ Generate cars$, XOR)$, $A_4 = (4, Softgoal,$ Easy to use$, AND)$ and $A_{13}=(13,$ $ArgElem,$ Redundant$)$. Note that the two arguments $A_5$ and $A'_5$ show an important difference between RationalGRL models and valid GRL models: While a valid GRL model disallows multiple elements with the same identifier (Definition~\ref{def:valid-grl-model}, condition 1), RationalGRL models do not enforce this restriction. This is because it is possible to create multiple arguments for the same element, where each argument contains different content for the same element. However, the set of \emph{accepted} elements in a RationalGRL should all have unique identifier (see Definition~\ref{def:valid-rationalgrl-model}).

@@

As an example, take the arguments $A_5$ for the goal `Generate cars' ($AND$ decomposition), $A'_5$ for the goal `Generate cars' ($XOR$ decomposition), $A_4$ for the softgoal `Easy to use' and the generic argument $A_{13}$ (`Redundant') against the softgoal (`Easy to use'). Given these arguments there are two attack links (Figure~\ref{fig:example-small3}), namely $(A'_{5},A_{5})$ and $(A_{13}, A_{4})$.

@@

\begin{flalign*}
A_0 = &(0, Actor, \text{Traffic simulator})&\\
A_1 = &(1, Task, \text{Car objects}, AND),&\\
A_2 = &(2, Softgoal, \text{Realistic simulation}, AND),&\\
A_3 = &(3, Softgoal, \text{Simple design}, AND),&\\
A_4 = &(4, Softgoal, \text{Easy to use}, AND),&\\
A_5 = &(5, Goal, \text{Generate cars}, AND),&\\
A'_5 = &(5, Goal, \text{Generate cars}, XOR),&\\
A_6 = &(6, Task, \text{Create new cars}, AND),&\\
A_7 = &(7, Task, \text{Wrap cars}, AND),&\\
A'_7 = &(7, Task, \text{Keep same cars}, AND),&\\
A_8 = &(8, Dep, 4, 1),&\\
A_9 = &(9, PosContr, 6, 2),&\\
A_{10} = &(10, NegContr, 6, 3),&\\
A_{11} = &(11, Decomp, 6, 5),&\\
A_{12} = &(12, Decomp, 7, 5),&\\
A_{13} = &(13, GenArg, \text{Redundant}),&\\
A_{14} = &(14, GenArg, \text{Necessary}),&\\
A_{15} = &(15, GenArg, \text{No effect}),&\\
A_{16} = &(16, GenArg, \text{Changes traffic flow}),&\\
A_{17} = &(ActIE,0,1),\ldots,A_{23}=(ActIE,0,7)&\\
\end{flalign*}
This model is then formalized as $RatGRL=(Arg, Att)$ where:
\begin{flalign*}
Arg = &\{A_0,\ldots,A_5, A'_5,A_6,A_7,A'_7\ldots,A_{23}\}\\
Att= &\{(A'_{5},A_5), (A'_{7},A_{7}), (A_{13},A_{4}), (A_{14},A_{13}),\\
     & (A_{15},A_{9}),(A_{16},A_{15})\}\\
\end{flalign*}

@@

As an example, suppose we are adding the decomposition of goal ``Generate cars'', expressed as argument $A_5 = (5,$ $Goal,$ Generate cars$, AND)$, into task ``Keep same cars'', and suppose we have the following argument for the goal: . Adding the decomposition can be formalized as executing $AS_5(5, \text{Keep same cars})$, and results in two new arguments: $A_7 = (7, Task, \text{Keep same cars}, AND)$ and $A_{12} = (12, Decomp, 7, 5)$, which are both added to the set of arguments $Arg$.

### 2017-11-07 introduction.tex
small fixes

The first challenge with the goal modeling phase, particularly in large projects, is related to their dynamic nature: goal models continuously change and evolve. Stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Stakeholders may change their opinions about a modeling decision in between two modeling sessions, which may require revisions of a goal model. If the rationales behind these revisions are not properly documented, alternative ideas and opposing views that could potentially lead to different goal models might be lost, as the resulting goal model only shows the end product of a long process and not the discussions during the modeling process. Furthermore, other stakeholders, such as developers who were not the original authors of the goal model, may have to make sense of a goal model in order to, for example, use it as an input in a later RE stage or in the development phase. If preferences, opinions and rationales behind the goal models are not stored explicitly, it may not only be more difficult to understand the model, but the other stakeholders may end up having similar discussions throughout the design and development phase as well.

Another challenge is that current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what its effect is on a goal model. Similarly, with existing goal modeling languages one can change a part of a goal model without being able to reason about whether or not this new goal model is consistent with the underlying beliefs and arguments. This becomes even more challenging if the stakeholders constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs of another group of stakeholders. The disconnect between goal models and their underlying beliefs and opinions may further lead to a poor understanding of the problem and solution domain, which is an important reason of RE project failure~\cite{curtis1988field}.

@@

\begin{enumerate}
\item 
The argumentation techniques must capture the actual discussions of the stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and their underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the underlying argumentation on the goal model, and vice versa.
\item 
A methodology must be identified that can guide practitioners in using the framework.
\item 
The framework must have a tool support.
\end{enumerate}

@@

Following from our previous work~\cite{vanzee-etal:renext2015,vanZee-etal:er2016}, we develop the \emph{RationalGRL} framework, which extends an existing framework for goal modeling, the Goal-oriented Requirements Language (GRL) \cite{Amyot:2010:EGM:1841349.1841356}, with a technique from argumentation theory called \emph{argument schemes} (or argumentation schemes~\cite{walton-etal2008}). Argument schemes are reusable patterns of reasoning that capture the common approaches in which humans argue and reason. We develop a list of argument schemes by performing an extensive case study in which we analyzed a set of transcripts containing more than 4 hours of discussions among designers of a traffic simulator information system. In order to specify clearly in what way RationalGRL extends GRL, we develop a metamodel, which specifies the traceability links between the arguments based on the schemes and the GRL models. In addition to this meta-model, we provide formal semantics for RationalGRL by formalizing the GRL language in propositional logic and rendering arguments about a GRL model as a formal argumentation framework~\cite{Dung1995}. We, then, formally capture the link between argumentation and goal modeling as a set of algorithms for applying argument schemes and critical questions about goal models.

In order to support practitioners in using the RationalGRL framework, we propose a methodology, which consists of developing goal models and posing arguments based on schemes in an integrated way. We  implemented a web-based tool for the RationalGRL framework as well, which works on any modern browser and is developed in Javascript.\footnote{The tool and source code can be found at \rationalgrlurl.} The tool is open-source, and designed such that new argument schemes and critical questions can be added easily.

>>>>

The first challenge for the goal modeling phase, particularly in large projects, is related to its dynamic nature: goal models continuously change and evolve. Stakeholders' preferences are rarely absolute, relevant, stable, or consistent~\cite{march1978bounded}. Stakeholders may change their opinions about a modeling decision in between two modeling sessions, which may require revisions of a goal model. If the rationales behind these revisions are not properly documented, alternative ideas and opposing views that could potentially lead to different goal models might be lost, as the resulting goal model only shows the end product of a long process and not the discussions during the modeling process. Furthermore, other stakeholders, such as developers who were not the original authors of the goal model, may have to make sense of a goal model in order to, for example, use it as an input in a later RE or development phase. If the preferences, opinions and rationales behind the goal models are not stored explicitly, it may not only be more difficult to understand the model, but the other stakeholders may end up having similar discussions throughout the design and development phase as well.

Another challenge is that current goal modeling languages have limited support for reasoning about changing beliefs and opinions, and their effect on the goal model. A stakeholder may change his or her opinion, but it is not always directly clear what the effect of this is on a goal model. Similarly, with existing goal modeling languages one can change a part of a goal model without being able to reason about whether or not this new goal model is consistent with the underlying beliefs and arguments. This becomes even more challenging if the stakeholders constructing the goal model change, since modeling decisions made by one group of stakeholders may conflict with the underlying beliefs of another group of stakeholders. The disconnect between goal models and their underlying beliefs and opinions may further lead to a poor understanding of the problem and solution domain, which is an important reason of RE project failure~\cite{curtis1988field}.

@@

\begin{enumerate}
\item 
The argumentation techniques must capture the actual discussions of the stakeholders or designers in the early requirements engineering phase.
\item 
The framework must have formal traceability links between elements of the goal model and their underlying arguments.
\item 
Using these traceability links, it must be possible to compute the effect of changes in the underlying argumentation on the goal model, and vice versa.
\item 
A methodology must be identified that can guide practitioners in using the framework.
\item 
The framework must have tool support. %"a tool support" is incorrect
\end{enumerate}

@@

Following from our previous work~\cite{vanzee-etal:renext2015,vanZee-etal:er2016}, we develop the \emph{RationalGRL} framework, which extends an existing framework for goal modeling, the Goal-oriented Requirements Language (GRL) \cite{Amyot:2010:EGM:1841349.1841356}, with a technique from argumentation theory called \emph{argument schemes} (or argumentation schemes~\cite{walton-etal2008}). Argument schemes are reusable patterns of reasoning that capture the typical ways in which humans argue and reason. Associated with argumentation schemes are so-called \emph{critical questions}, which can point to typical sources of doubt or implicit assumptions people make when arguing in a particular way. We believe that argument schemes are are a good choice for modeling discussions about a goal model, as they can guide users in systematically deriving conclusions and making assumptions explicit~\cite{bexEtal2003,murukannaiah2015}.

Inspired by the work on practical reasoning from Artificial Intelligence, most notably Atkinson and Bench-Capon~\cite{atkinson2007}, we have developed a list of argument schemes that can be used to analyze and guide stakeholders' discussions about goal models. This list of argument schemes is based on an extensive case study in which we analyzed a set of transcripts containing more than 4 hours of discussions among designers of a traffic simulator information system.

In order to specify clearly in what way RationalGRL extends GRL, we develop a metamodel, which specifies the traceability links between the arguments based on the schemes and the GRL models. In addition to this meta-model, we provide formal semantics for RationalGRL by formalizing the GRL language in propositional logic and rendering arguments about a GRL model as a formal argumentation framework~\cite{Dung1995}. We, then, formally capture the link between argumentation and goal modeling as a set of algorithms for applying argument schemes and critical questions about goal models.

In order to support practitioners in using the RationalGRL framework, we propose a methodology, which consists of developing goal models and posing arguments based on schemes in an integrated way. We implement a web-based tool for the RationalGRL framework as well, which works on any modern browser and is developed in Javascript.\footnote{The tool and source code can be found at \rationalgrlurl.} The tool is open-source, and designed such that new argument schemes and critical questions can be added easily.

### 2017-11-07 methodology.tex
small fixes

The final requirement of our framework is that it has tool support (\textbf{requirement 5}). This is important for various reasons: i) although there are various approaches attempting to combine goal modeling with argumentation, we are not aware of integrated tool support (see Section~\ref{sect:discussion}), ii) it allows us to do user tests in future research, exploring the difference between GRL and RationalGRL, iii) tool support is an excellent verification to ensure our formal framework and algorithms actually work. In this section we briefly highlight some features of the tool, and we explain some of its current limitations.

@@

As we discussed in Section~\ref{sect:background:grl}, GRL has a well-documented and well-maintained tool called jUCMNav~\cite{jUCMNav}\footnote{\url{http://jucmnav.softwareengineering.ca/foswiki/ProjetSEG}}, which is implemented as an Eclipse plugin. Since this tool is open-source, many researchers have chosen to extend it in various way. However, our tool is not yet integrated into jUCMNav, but consists of a stand-alone web-based version. While we aim to integrate our framework into jUCMNav in the future, a web-based version of RationalGRL (and GRL) is useful by itself as well. The authors of this article have been involved in a large number of empirical studies involving jUCMNav, and our experience is that it can take significant time for participants to setup the Eclipse environment and install the jUCMNav plugin properly. Since our extension does not make use of the many features of jUCMNav, we believe a light-weight web-based version is useful for future empirical research, and it may benefit the community in general. Since RationalGRL is an extension of GRL, our web-based version allows for (limited) GRL modeling. The tool can be accessed from:

>>>>

The final requirement of our framework is that it has tool support (\textbf{requirement 5}). This is important for various reasons: i) although there are various approaches attempting to combine goal modeling with argumentation, we are not aware of integrated tool support (see Section~\ref{sect:discussion}), ii) it allows us to do user tests in future research, exploring the difference between GRL and RationalGRL, iii) tool support is an excellent verification to ensure our formal framework and algorithms actually work. In this section we briefly highlight some features of the tool, and we explain some of its current limitations. The tool can be accessed from:

@@

GRL is implemented in the open-source tool jUCMNav~\cite{jUCMNav}, which is actively maintained\footnote{\url{http://jucmnav.softwareengineering.ca/foswiki/ProjetSEG}}. jUCMNav currently has over 2.000 commits made by over 40 contributors, representing over 200.000 lines of code. jUCMNav is used actively in research, and the tool has been extended in many directions over the past few years, such as for modeling and analyzing security misuse cases \cite{}, supporting activity theory \cite{}, Web-based goal modeling \cite{}, and enterprise architecture modeling \cite{}. \todo{Floris}{Marc}{references need to be added here!} Furthermore, jUCMNav includes Use Case modelling and contains various GRL evaluation algorithms with which the satisfiability of goals can be determined.

The jUCMNav tool is implemented as an Eclipse plugin. While we aim to integrate our framework into jUCMNav in the future, a web-based version of RationalGRL (and GRL) is more suited for current purposes. For example, it is our experience is that it can take significant time to set up the Eclipse environment and install the jUCMNav plugin, which hampers quick uptake by practitioners and also makes setting up empirical evaluations more difficult. Since RationalGRL does not make use of the many features of jUCMNav, we believe a light-weight version is sufficient for our current aims, and may benefit the community in general.
